Do syntactic and semantic similarity lead to interference effects?\\ Evidence from self-paced reading and event-related potentials using German

### 2024-11-01 paper/revision2/elsarticle-template-harv.tex
Add files via upload

\documentclass[a4paper, man, floatsintext]{apa7}

% Language setting
\usepackage[english]{babel}

% Useful packages
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
%\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage[natbib=true,maxcitenames=2,bibstyle=authoryear, style=apa]{biblatex}
\usepackage{csquotes}
\DeclareLanguageMapping{american}{american-apa}
\bibliography{interferencebib}

\usepackage{gb4e}
\usepackage{pifont}
\noautomath
\usepackage{authblk}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{subcaption}

\usepackage{bigdelim}

\usepackage{clipboard}
\newclipboard{output-reviews}
\openclipboard{output-reviews}

\title{\hspace{1cm} Do syntactic and semantic similarity lead to interference effects?
\newline
Evidence from self-paced reading and event-related potentials using German}

\authorsnames[1,{2},1]{Pia Schoknecht, Himanshu Yadav, Shravan Vasishth}
\authorsaffiliations{{Department of Linguistics, University of Potsdam},
{Department of Cognitive Science, Indian Institute of Technology Kanpur}}

\shorttitle{Syntactic and semantic interference effects}

\authornote{This research was funded by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation), Project ID 317633480, SFB 1287.
We thank Johanna Thieke, Romy Leue, Elise Oltrogge and Lisa Plagemann for assistance during data acquisition. We are grateful to Dario Paape for helpful discussions.
Correspondence concerning this article should be addressed to Pia Schoknecht, e-mail: pia.schoknecht@uni-potsdam.de
}

\abstract{Cue-based retrieval accounts of sentence processing postulate that at a verb, retrieval cues are generated to complete a dependency with the verb's argument(s); for example, the dependency between the subject and the verb must be completed to interpret the sentence. If these retrieval cues match with not only the features on the noun that is the retrieval target but also with those on other nouns in the sentence, then processing difficulty arises at the verb. This difficulty in identifying the correct dependent is called similarity-based interference. 
We present large-sample self-paced reading (N = 774) and event-related potentials experiments (N = 103) using a well-established design \citep{vandyke07,mertzen}, to investigate interference due to syntactic and semantic cues in German. 
A Bayes factor analysis showed that both the self-paced reading times as well as event-related potentials provided clear evidence for a semantic interference effect. In the self-paced reading experiment, interference due to the semantic cues caused a reading time slow-down starting at the distractor. 
The event-related potentials showed a more negative N400 amplitude at the critical verb, i.e., the retrieval site, for high vs.\ low semantic interference.
Surprisingly, in both experiments, Bayes factor analyses showed evidence against interference due to syntactic cues. This finding contradicts the predictions of the standard implementations of cue-based retrieval theory, which (implicitly) assume that both syntactic and semantic cues play an equal role in retrieval. We show through computational modeling that cue-based retrieval also shows no syntactic interference if the parser is assumed to keep track of which clause the subject occurs in. Thus, if syntactic retrieval cues include hierarchical syntactic information (is the noun in the same clause as the verb?), the predictions of the cue-based retrieval model would be consistent with the observed patterns in our data.

In sum, this large-sample study of German using self-paced reading and event-related potentials suggests that, in our data, although both syntactic and semantic cues drive retrieval at the verb, syntactic cues may include hierarchical syntactic information, which can render the parser immune to syntactic interference effects.

\clearpage

}

@@

\maketitle

@@

\Copy{intro}{

It is well-established in sentence processing that completing linguistic dependencies requires memory retrieval \textcolor{blue}{\parencite[see e.g., ][]{lewis06}. In fact, a key assumption in psycholinguistics is that the same constraints on memory retrieval that have been identified within memory research in cognitive psychology \parencite[e.g., ][]{anderson_etal_2004_ACT_R} may be applicable in sentence processing as well. An example of a key construct from cognitive psychology that is relevant for sentence processing is that spreading activation leads to difficulty in identifying the correct item to retrieve from memory. The present work investigates whether and how such independently posited, general constraints on memory retrieval impact sentence processing. As a concrete example of general memory constraints impacting sentence comprehension,} consider (\ref{ex:1_vandyke07}).

\begin{exe}[ht]
\ex \label{ex:1_vandyke07} 
The worker was surprised that the resident who was living near the dangerous warehouse was complaining about the investigation. \parencite{vandyke07}
\end{exe}

There is clear evidence in the psycholinguistic literature \parencite[e.g.,][]{jaeger_etal_2020, nicenboim, vandyke_mcelree06, vandyke_lewis03} that, at the verb \textit{was complaining}, it is necessary to retrieve the subject \textit{the resident} from memory in order to interpret who did what to whom. A widely-held assumption is that such memory retrievals during sentence comprehension are guided by retrieval cues \textcolor{blue}{\citep{Lewis2005, mcelree}}.}\label{intro}

Here, we consider two kinds of cues used in retrieval: semantic and  syntactic cues.
In a sentence like (\ref{ex:1_vandyke07}), the retrieval cues generated at the verb are standardly assumed \parencite[e.g.,][]{vandyke07} to include at least the semantic cue \{+animate\} and the syntactic cue \{+grammatical subject\}. These cues pick out the target noun \textit{the resident} as the subject. The distractor \textit{warehouse}, intervening between subject and verb, does not match either of these retrieval cues. However, if the distractor were \textit{neighbor} instead of \textit{warehouse}, then it would match the \{+animate\} cue at the verb. This overlap in cues between the target noun and the distractor is called cue-overload in cue-based retrieval accounts, and is predicted to trigger semantic interference, which expresses itself as greater processing time at the verb compared to the baseline condition (\ref{ex:1_vandyke07}).

@@

Previous work has shown clear evidence for this kind of retrieval interference \parencite{vandyke07, vandyke_mcelree06, vandyke_mcelree2011, nicenboim,  vandyke_lewis03,jaeger_etal_2017}. However, two important questions remain unanswered. First, as discussed in \textcite{jaeger_etal_2017}, almost all the published studies on interference effects are severely underpowered. Low power has the consequence that published estimates will tend to be overestimates or misestimates of the effect size \parencite{vasishth2018_signficancefilter}. Over-/misestimates have important implications for theory evaluation. One example is discussed in \textcite{vasishth2018_signficancefilter}, where a key prediction of surprisal theory \parencite{levy&keller_2013} could not be validated when a larger-sample study was conducted. By contrast, the original small-sample study \parencite{levy&keller_2013} showed suggestive evidence consistent with the theory. Another example is the recent heated debate \parencite{nieuwland_etal_2018} on prediction in sentence comprehension, which was largely focused on the results of low-powered studies, leading to potentially misleading conclusions \parencite{nicenboim_etal_2020}.
Second, although it has been established \parencite[e.g.,][]{vandyke07,mertzen} that syntactic and semantic cues are involved in sentence processing, it is not obvious which cues are relevant for a particular retrieval event. For example, \textcite{dillon2013} argued, following \textcite{Sturt2003}, that the gender cue was either downweighted or not used at all when an antecedent-reflexive dependency needs to be built. Instead, the antecedent of a reflexive could be identified using the syntactic principle A of the binding theory, which correctly identifies the target for retrieval.  Using computational modeling, \textcite{dillon2013} suggested that antecedent-reflexive dependencies were immune to interference, and that the reason for this is that principle A of the binding theory is the primary driver of the cue-based retrieval process \parencite[cf. ][]{jaeger_etal_2020,yadav2021individual}. This was a surprising discovery because the default assumption in cue-based retrieval accounts was that all cues were equally weighted, regardless of the dependency type; this had the consequence that retrieval interference was expected whenever a cue-overload configuration was present in a sentence. More broadly, an important insight in the work of \textcite{dillon2013}, among others, is that  the precise details of which cues are used during retrieval can affect processing difficulty in possibly surprising ways. We build on this idea in the present paper.

@@

Reading studies have been the dominant method used to investigate retrieval interference \parencite[see the review in][]{jaeger_etal_2017}. We aimed to investigate the role of syntactic and semantic cues during sentence comprehension with higher-powered experiments using  self-paced reading (SPR) as well as electroencephalography (EEG), which to our knowledge has never been used to investigate interference during subject-verb dependency resolution.

\Copy{vd_design}{All experiments used the 2 $\times$ 2 design developed by \textcite{vandyke07}; the design crosses syntactic and semantic interference.} \label{vd_design}

@@

Syntactic interference is manipulated via the syntactic status of the distractor, which is either the subject (\ref{ex:hisyn}) of an embedded clause, or part of a prepositional phrase (\ref{ex:losyn}) and hence a non-subject. \Copy{franck_refs}{Following \textcite{vandyke07}, the assumption here is that  during the retrieval of the subject, a distractor in \emph{any} subject position causes more syntactic interference than a distractor inside a prepositional phrase \textcolor{blue}{ -- even if the distractor occurs in an embedded clause {\parencite[this assumption stands in contrast to findings on so-called agreement attraction, where syntactic hierarchy plays a crucial role, see e.g.,][]{franck2002subject, franck2006agreement,franck2011_commentary, franck2020hierarchical}.}}}\label{franck_refs}

Semantic interference is manipulated via the animacy of the distractor. During the retrieval of the subject of the verb \textit{trank} (`drank'), the animate distractor  \textit{Einbrecher} (`burglar') is assumed to cause high semantic interference, while the inanimate distractor \textit{Verlust} (`loss') is assumed to cause low semantic interference. \Copy{semantic_features_encoding}{\textcolor{blue}{The animacy of the distractor is used as a stand-in feature for semantic features relevant for the specific retrieval event which might be more fine-grained. In \ref{ex:materials}, the semantic retrieval cue might be \{+animate\} or \{+human\} but it might as well be \{+can drink\}. While more generic features like \{+animate\} or \{+human\} (all animate distractors were human) might be created during memory encoding, more specific features like \{+can drink\} are probably only activated when relevant. The present study is not designed to provide insight into the specific semantic features used during memory encoding and retrieval. For experimental work on highly specific retrieval cues like \{+sailable\} and \{+shatterable\}, see \textcite{vandyke_mcelree06} and \textcite{cunnings_sturt_2018retrieval}. For a principled approach to feature selection using word embeddings, see \textcite{smith2020principled}. A plausibility norming experiment (see page \pageref{plausib_normin_orig}) was conducted to ensure that the semantic manipulation did work, i.e., that the semantic requirements of the verb matched the features of the subject and the animate distractor, but not the features of the inanimate distractor.
}}\label{semantic_features_encoding}

\paragraph{Regions of interest in the design}
\textcolor{blue}{For both our SPR and EEG experiments, o}ur principal focus was on the processing of the verb (\textit{trank}, `drank'). \textcolor{blue}{In the SPR experiment, t}he reading times of the immediately following spill-over region and of the pre-critical region were also of interest. Previous work using the present design \parencite{vandyke07,mertzen} has consistently found effects in the pre-critical region (we return to this point in the discussion sections\textcolor{blue}{, see Figure \ref{fig:previous_vs_present} A}). A further issue with German embedded clauses \parencite[also see][]{mertzen} was that, without an appropriate intervening phrase, the clause embedding would lead to two verbs appearing one after another (\textit{erzählt hatte, trank}, literally `told, drank'). This is because in German, the verb is clause-final in embedded clauses. Such verb sequences are inherently difficult to process \parencite{VSLK11,bach1986crossed}.

For these reasons, we added two adverbs before the critical verb. Including these intervening regions had a further advantage: 
It is well-known \parencite[e.g., ][]{rayner2000effect} that the last word in a clause-final position causes a slowdown in reading; such a slowdown has the potential to spill over to the subsequent regions, possibly affecting the estimated effects at the critical region. The spill-over of longer reading times can be problematic because the standard deviation increases as well \parencite{wagenmakers2007linear}; such an increase in standard deviation would drastically lower statistical power. 
Inserting two adverbs before the critical verb thus served to attenuate spill-over from the clause-final word in the embedded clause preceding the critical region. This way, the critical region should be less affected by the clause-boundary slowdown. This design also has the advantage that any potential effects seen at the pre-critical region (the adverb immediately preceding the verb) are less likely to be due to spill-over from the preceding region. Thus, our focus in the \textcolor{blue}{SPR experiment was} on the processing difficulty observed at the critical verb and the surrounding regions.

\Copy{why_n400}{\textcolor{blue}{In the EEG experiment, the region of interest was the critical verb. Due to the direct measurement of brain activity with excellent temporal resolution event-related potential (ERP) effects do not tend to spill-over into later regions. We focused specifically on the N400 elicited by the verb. The N400 is a negative deflection in the ERP signal that peaks approximately 400 ms after any word was encountered. Traditionally, it is associated with lexical-semantic processing, but the finding of multi-modal N400 effects suggests that it rather indexes differences in the processing of meaningful stimuli in general \citep{kutas_federmeier2011}. The N400 amplitude is greater for words that deviate from the meaning activated by its context compared to words that fit its context \citep[see e.g.,][]{kutas_hillyard, Kutas_Iragui_1998}. Consequently, the amplitude of the N400 to a specific word is a negative function of its predictability within the given context \citep[see e.g.,][]{kutas_hillyard1984, frank2015_erp_surprisal}. Additionally, the N400 elicited by isolated words is negatively correlated with their frequency \citep{kutas&federmeier_2000}. Therefore, the N400 amplitude is thought to index a) the increased difficulty to integrate a semantically anomalous or simply unexpected word into the sentence context \citep[e.g.,][]{hagoort2004integration} and/or b) the necessity to retrieve lexical information from memory which was not already pre-activated before the stimulus was perceived \citep[e.g.,][]{brouwer2017_n4_p6}. These two accounts of the N400 have been extensively discussed in the ERP literature on prediction \citep[see e.g.,][]{nieuwland_etal_2018, nicenboim_etal_2020, Freunberger_Roehm_2017, Delong_etal_2005, mantegna2019, nieuwland2019_integrationprediction}. As a likely marker of memory retrieval during sentence comprehension, it is fair to assume that the N400 would show modulations due to retrieval interference. The few ERP studies that have investigated retrieval interference so far have indeed found negativities \citep{lee_garnsey, martinetal2014, schoknecht2022, vasishth_drenhaus_2011}, although the topography and latency of these negativities was mostly not typical for the N400. It is important to highlight here that the previous ERP studies on interference either increased the number of distractors in a sentence or used grammatical gender or number as the critical retrieval cue. So, interference due to non-semantic features elicited negativities which were more or less similiar to the N400. Therefore, we expected the interference manipulation regardless of the type of retrieval cue (syntactic or semantic) to elicit N400 or similar effects. For a discussion of the relation between the N400 and other negativities, see \textcite{ina_n400}.}

\textcolor{blue}{Alternately, retrieval interference -- especially syntactic retrieval interference -- might modulate the P600. The P600 is a positive deflection in the ERP wave form that is typically observed at least 600 ms after stimulus onset. It is associated with syntactic processing, e.g., morpho-syntactic errors and reanalysis \citep[see, e.g.,][]{osterhout&holcomb_1992, kaan_etal_2000_P600}. High interference could increase difficulty during syntactic integration of subject and verb in our materials. Potential misretrievals of the distractor might lead to reanalysis which could manifest as an increased P600 amplitude in high compared to low interference conditions.}} \label{why_n400}

@@

\paragraph{Procedure for the plausibility study}
The plausibility norming experiment was implemented in PCIbex \citep{pcibex} and was conducted online. Participants were asked to judge the plausibility of the sentences on a 7-point scale ranging from absolutely implausible (1) to absolutely plausible (7). All sentences belonging to one item were shown on one screen, so that they could be judged relative to each other. The plausibility norming experiment included 120 sentences from a different experiment which functioned as fillers. The participants received \pounds 3.40 as compensation for the norming experiment. A demonstration of the plausibility norming experiment can be accessed via \hyperlink{https://farm.pcibex.net/r/WIeKYM/}{https://farm.pcibex.net/r/WIeKYM/}.

@@

\begin{figure}[H]
\caption{Plausibility ratings for the animate (high semantic interference conditions) and inanimate distractor (low semantic interference conditions), the introduction noun (\textit{X believed that \dots}) and the subject. The diamonds represent the mean rating. Blue dots show the individual data points. }\label{fig:plausibility_rating}
\centering
\includegraphics[width=0.6\textwidth]{images/pandora_plausibility_ratings.jpg}
\end{figure}

@@

As discussed in detail below, the second session showed an adaptation effect \parencite{prasad2021rapid}: Reading times decreased overall. The adaptation effect is theoretically interesting per se \parencite{fine2013rapid} and we discuss it below; but it was problematic for our design because the average effect would then be confounded by adaptation. For this reason, we ran a second version of the SPR study, hereafter Experiment 1b, in which participants (N=570) were either shown items 1-60 or items 61-120 (but not both). In the Results section, we present both sets of results as well as the pooled data (N=774) from E1a and E1b.
As mentioned above, because the ERP method requires a large number of items, all the ERP data come from participants who completed both sessions.

@@

\begin{figure}[H]
    \caption{The structure of the present study. In total, 120 items were used for the SPR (red branches) and the ERP experiments (blue branches). The color brightness reflects how many sessions were completed by each participant (lighter color: two sessions, darker color: one session).}
    \label{fig:project_str}
    \centering
    \includegraphics[width=\textwidth]{images/pandora_project_structure_figure.pdf}
\end{figure}

>>>>

%% 
%% Copyright 2007-2020 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%% 
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%% 
%% Template article for Elsevier's document class `elsarticle'
%% with harvard style bibliographic references

\documentclass[review,preprint,12pt,authoryear,floatsintext]{elsarticle}

% my packages
\usepackage[english]{babel}

\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{rotating}
\usepackage{subfig}
\usepackage{color}
\usepackage{gb4e}
\usepackage{pifont}
\noautomath
\usepackage{amsmath} 
\usepackage{rotating}
\usepackage{multirow}
\usepackage{subcaption}

\usepackage{bigdelim}
\usepackage{hyperref}

%my packages end

%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times,authoryear]{elsarticle}
%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,3p,times,authoryear]{elsarticle}
%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,5p,times,authoryear]{elsarticle}
%% \documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}

%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
%% please give \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
%% \usepackage{lineno}

@@

\begin{frontmatter}
\title{Do syntactic and semantic similarity lead to interference effects?\\
Evidence from self-paced reading and event-related potentials using German}
%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \affiliation for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \affiliation{organization={},
%%            addressline={}, 
%%            city={},
%%            postcode={}, 
%%            state={},
%%            country={}}
%% \fntext[label3]{}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{}
%% \affiliation[label1]{organization={},
%%             addressline={},
%%             city={},
%%             postcode={},
%%             state={},
%%             country={}}
%%
%% \affiliation[label2]{organization={},
%%             addressline={},
%%             city={},
%%             postcode={},
%%             state={},
%%             country={}}

\begin{abstract}
%% Text of abstract
Cue-based retrieval accounts of sentence processing postulate that at a verb, retrieval cues are generated to complete a dependency with the verb's argument(s); for example, the dependency between the subject and the verb must be completed to interpret the sentence. If these retrieval cues match with not only the features on the noun that is the retrieval target but also with those on other nouns in the sentence, then processing difficulty arises at the verb. This difficulty in identifying the correct dependent is called similarity-based interference. 
We present large-sample self-paced reading (N = 774) and event-related potentials experiments (N = 103) using a well-established design \citep{vandyke07,mertzen}, to investigate interference due to syntactic and semantic cues in German. 
A Bayes factor analysis showed that both the self-paced reading times as well as event-related potentials provided clear evidence for a semantic interference effect. In the self-paced reading experiment, interference due to the semantic cues caused a reading time slow-down starting at the distractor. 
The event-related potentials showed a more negative N400 amplitude at the critical verb, i.e., the retrieval site, for high vs.\ low semantic interference.
Surprisingly, in both experiments, Bayes factor analyses showed evidence against interference due to syntactic cues. This finding contradicts the predictions of the standard implementations of cue-based retrieval theory, which (implicitly) assume that both syntactic and semantic cues play an equal role in retrieval. We show through computational modeling that cue-based retrieval also shows no syntactic interference if the parser is assumed to keep track of which clause the subject occurs in. Thus, if syntactic retrieval cues include hierarchical syntactic information (is the noun in the same clause as the verb?), the predictions of the cue-based retrieval model would be consistent with the observed patterns in our data.

In sum, this large-sample study of German using self-paced reading and event-related potentials suggests that, in our data, although both syntactic and semantic cues drive retrieval at the verb, syntactic cues may include hierarchical syntactic information, which can render the parser immune to syntactic interference effects.
\end{abstract}

\end{frontmatter}
%% \linenumbers

%% main text
\newpage

@@

{

It is well-established in sentence processing that completing linguistic dependencies requires memory retrieval \citep[see e.g., ][]{lewis06}. In fact, a key assumption in psycholinguistics is that the same constraints on memory retrieval that have been identified within memory research in cognitive psychology \citep[e.g., ][]{anderson_etal_2004_ACT_R} may be applicable in sentence processing as well. An example of a key construct from cognitive psychology that is relevant for sentence processing is that spreading activation leads to difficulty in identifying the correct item to retrieve from memory. The present work investigates whether and how such independently posited, general constraints on memory retrieval impact sentence processing. As a concrete example of general memory constraints impacting sentence comprehension, consider (\ref{ex:1_vandyke07}).

\begin{exe}[ht]
\ex \label{ex:1_vandyke07} 
The worker was surprised that the resident who was living near the dangerous warehouse was complaining about the investigation. \citep{vandyke07}
\end{exe}

There is clear evidence in the psycholinguistic literature \citep[e.g.,][]{jaeger_etal_2020, nicenboim, vandyke_mcelree06, vandyke_lewis03} that, at the verb \textit{was complaining}, it is necessary to retrieve the subject \textit{the resident} from memory in order to interpret who did what to whom. A widely-held assumption is that such memory retrievals during sentence comprehension are guided by retrieval cues \citep{Lewis2005, mcelree}.}\label{intro}

Here, we consider two kinds of cues used in retrieval: semantic and  syntactic cues.
In a sentence like (\ref{ex:1_vandyke07}), the retrieval cues generated at the verb are standardly assumed \citep[e.g.,][]{vandyke07} to include at least the semantic cue \{+animate\} and the syntactic cue \{+grammatical subject\}. These cues pick out the target noun \textit{the resident} as the subject. The distractor \textit{warehouse}, intervening between subject and verb, does not match either of these retrieval cues. However, if the distractor were \textit{neighbor} instead of \textit{warehouse}, then it would match the \{+animate\} cue at the verb. This overlap in cues between the target noun and the distractor is called cue-overload in cue-based retrieval accounts, and is predicted to trigger semantic interference, which expresses itself as greater processing time at the verb compared to the baseline condition (\ref{ex:1_vandyke07}).

@@

Previous work has shown clear evidence for this kind of retrieval interference \citep{vandyke07, vandyke_mcelree06, vandyke_mcelree2011, nicenboim,  vandyke_lewis03,jaeger_etal_2017}. However, two important questions remain unanswered. First, as discussed in \cite{jaeger_etal_2017}, almost all the published studies on interference effects are severely underpowered. Low power has the consequence that published estimates will tend to be overestimates or misestimates of the effect size \citep{vasishth2018_signficancefilter}. Over-/misestimates have important implications for theory evaluation. One example is discussed in \cite{vasishth2018_signficancefilter}, where a key prediction of surprisal theory \citep{levy&keller_2013} could not be validated when a larger-sample study was conducted. By contrast, the original small-sample study \citep{levy&keller_2013} showed suggestive evidence consistent with the theory. Another example is the recent heated debate \citep{nieuwland_etal_2018} on prediction in sentence comprehension, which was largely focused on the results of low-powered studies, leading to potentially misleading conclusions \citep{nicenboim_etal_2020}.
Second, although it has been established \citep[e.g.,][]{vandyke07,mertzen} that syntactic and semantic cues are involved in sentence processing, it is not obvious which cues are relevant for a particular retrieval event. For example, \cite{dillon2013} argued, following \cite{Sturt2003}, that the gender cue was either downweighted or not used at all when an antecedent-reflexive dependency needs to be built. Instead, the antecedent of a reflexive could be identified using the syntactic principle A of the binding theory, which correctly identifies the target for retrieval.  Using computational modeling, \cite{dillon2013} suggested that antecedent-reflexive dependencies were immune to interference, and that the reason for this is that principle A of the binding theory is the primary driver of the cue-based retrieval process \citep[cf. ][]{jaeger_etal_2020,yadav2021individual}. This was a surprising discovery because the default assumption in cue-based retrieval accounts was that all cues were equally weighted, regardless of the dependency type; this had the consequence that retrieval interference was expected whenever a cue-overload configuration was present in a sentence. More broadly, an important insight in the work of \cite{dillon2013}, among others, is that  the precise details of which cues are used during retrieval can affect processing difficulty in possibly surprising ways. We build on this idea in the present paper.

@@

Reading studies have been the dominant method used to investigate retrieval interference \citep[see the review in][]{jaeger_etal_2017}. We aimed to investigate the role of syntactic and semantic cues during sentence comprehension with higher-powered experiments using  self-paced reading (SPR) as well as electroencephalography (EEG), which to our knowledge has never been used to investigate interference during subject-verb dependency resolution.

{All experiments used the 2 $\times$ 2 design developed by \cite{vandyke07}; the design crosses syntactic and semantic interference.} \label{vd_design}

@@

Syntactic interference is manipulated via the syntactic status of the distractor, which is either the subject (\ref{ex:hisyn}) of an embedded clause, or part of a prepositional phrase (\ref{ex:losyn}) and hence a non-subject. {Following \cite{vandyke07}, the assumption here is that  during the retrieval of the subject, a distractor in \emph{any} subject position causes more syntactic interference than a distractor inside a prepositional phrase -- even if the distractor occurs in an embedded clause {\citep[this assumption stands in contrast to findings on so-called agreement attraction, where syntactic hierarchy plays a crucial role, see e.g.,][]{franck2002subject, franck2006agreement,franck2011_commentary, franck2020hierarchical}.}}\label{franck_refs}

Semantic interference is manipulated via the animacy of the distractor. During the retrieval of the subject of the verb \textit{trank} (`drank'), the animate distractor  \textit{Einbrecher} (`burglar') is assumed to cause high semantic interference, while the inanimate distractor \textit{Verlust} (`loss') is assumed to cause low semantic interference. {The animacy of the distractor is used as a stand-in feature for semantic features relevant for the specific retrieval event which might be more fine-grained. In \ref{ex:materials}, the semantic retrieval cue might be \{+animate\} or \{+human\} but it might as well be \{+can drink\}. While more generic features like \{+animate\} or \{+human\} (all animate distractors were human) might be created during memory encoding, more specific features like \{+can drink\} are probably only activated when relevant. The present study is not designed to provide insight into the specific semantic features used during memory encoding and retrieval. For experimental work on highly specific retrieval cues like \{+sailable\} and \{+shatterable\}, see \cite{vandyke_mcelree06} and \cite{cunnings_sturt_2018retrieval}. For a principled approach to feature selection using word embeddings, see \cite{smith2020principled}. A plausibility norming experiment (see page \pageref{plausib_normin_orig}) was conducted to ensure that the semantic manipulation did work, i.e., that the semantic requirements of the verb matched the features of the subject and the animate distractor, but not the features of the inanimate distractor.
}\label{semantic_features_encoding}

\paragraph{Regions of interest in the design}
For both our SPR and EEG experiments, our principal focus was on the processing of the verb (\textit{trank}, `drank'). In the SPR experiment, the reading times of the immediately following spill-over region and of the pre-critical region were also of interest. Previous work using the present design \citep{vandyke07,mertzen} has consistently found effects in the pre-critical region (we return to this point in the discussion sections, see Figure \ref{fig:previous_vs_present} A). A further issue with German embedded clauses \citep[also see][]{mertzen} was that, without an appropriate intervening phrase, the clause embedding would lead to two verbs appearing one after another (\textit{erzählt hatte, trank}, literally `told, drank'). This is because in German, the verb is clause-final in embedded clauses. Such verb sequences are inherently difficult to process \citep{VSLK11,bach1986crossed}.

For these reasons, we added two adverbs before the critical verb. Including these intervening regions had a further advantage: 
It is well-known \citep[e.g., ][]{rayner2000effect} that the last word in a clause-final position causes a slowdown in reading; such a slowdown has the potential to spill over to the subsequent regions, possibly affecting the estimated effects at the critical region. The spill-over of longer reading times can be problematic because the standard deviation increases as well \citep{wagenmakers2007linear}; such an increase in standard deviation would drastically lower statistical power. 
Inserting two adverbs before the critical verb thus served to attenuate spill-over from the clause-final word in the embedded clause preceding the critical region. This way, the critical region should be less affected by the clause-boundary slowdown. This design also has the advantage that any potential effects seen at the pre-critical region (the adverb immediately preceding the verb) are less likely to be due to spill-over from the preceding region. Thus, our focus in the SPR experiment was on the processing difficulty observed at the critical verb and the surrounding regions.

{In the EEG experiment, the region of interest was the critical verb. Due to the direct measurement of brain activity with excellent temporal resolution event-related potential (ERP) effects do not tend to spill-over into later regions. We focused specifically on the N400 elicited by the verb. The N400 is a negative deflection in the ERP signal that peaks approximately 400 ms after any word was encountered. Traditionally, it is associated with lexical-semantic processing, but the finding of multi-modal N400 effects suggests that it rather indexes differences in the processing of meaningful stimuli in general \citep{kutas_federmeier2011}. The N400 amplitude is greater for words that deviate from the meaning activated by its context compared to words that fit its context \citep[see e.g.,][]{kutas_hillyard, Kutas_Iragui_1998}. Consequently, the amplitude of the N400 to a specific word is a negative function of its predictability within the given context \citep[see e.g.,][]{kutas_hillyard1984, frank2015_erp_surprisal}. Additionally, the N400 elicited by isolated words is negatively correlated with their frequency \citep{kutas&federmeier_2000}. Therefore, the N400 amplitude is thought to index a) the increased difficulty to integrate a semantically anomalous or simply unexpected word into the sentence context \citep[e.g.,][]{hagoort2004integration} and/or b) the necessity to retrieve lexical information from memory which was not already pre-activated before the stimulus was perceived \citep[e.g.,][]{brouwer2017_n4_p6}. These two accounts of the N400 have been extensively discussed in the ERP literature on prediction \citep[see e.g.,][]{nieuwland_etal_2018, nicenboim_etal_2020, Freunberger_Roehm_2017, Delong_etal_2005, mantegna2019, nieuwland2019_integrationprediction}. As a likely marker of memory retrieval during sentence comprehension, it is fair to assume that the N400 would show modulations due to retrieval interference. The few ERP studies that have investigated retrieval interference so far have indeed found negativities \citep{lee_garnsey, martinetal2014, schoknecht2022, vasishth_drenhaus_2011}, although the topography and latency of these negativities was mostly not typical for the N400. It is important to highlight here that the previous ERP studies on interference either increased the number of distractors in a sentence or used grammatical gender or number as the critical retrieval cue. So, interference due to non-semantic features elicited negativities which were more or less similiar to the N400. Therefore, we expected the interference manipulation regardless of the type of retrieval cue (syntactic or semantic) to elicit N400 or similar effects. For a discussion of the relation between the N400 and other negativities, see \cite{ina_n400}.

Alternately, retrieval interference -- especially syntactic retrieval interference -- might modulate the P600. The P600 is a positive deflection in the ERP wave form that is typically observed at least 600 ms after stimulus onset. It is associated with syntactic processing, e.g., morpho-syntactic errors and reanalysis \citep[see, e.g.,][]{osterhout&holcomb_1992, kaan_etal_2000_P600}. High interference could increase difficulty during syntactic integration of subject and verb in our materials. Potential misretrievals of the distractor might lead to reanalysis which could manifest as an increased P600 amplitude in high compared to low interference conditions.} \label{why_n400}

@@

\paragraph{Procedure for the plausibility study}
The plausibility norming experiment was implemented in PCIbex \citep{pcibex} and was conducted online. Participants were asked to judge the plausibility of the sentences on a 7-point scale ranging from absolutely implausible (1) to absolutely plausible (7). All sentences belonging to one item were shown on one screen, so that they could be judged relative to each other. The plausibility norming experiment included 120 sentences from a different experiment which functioned as fillers. The participants received \pounds 3.40 as compensation for the norming experiment. A demonstration of the plausibility norming experiment can be accessed via \\\hyperlink{https://farm.pcibex.net/r/WIeKYM/}{https://farm.pcibex.net/r/WIeKYM/}.

@@

\begin{figure}[ht]
\caption{Plausibility ratings for the animate (high semantic interference conditions) and inanimate distractor (low semantic interference conditions), the introduction noun (\textit{X believed that \dots}) and the subject. The diamonds represent the mean rating. Blue dots show the individual data points. }\label{fig:plausibility_rating}
\centering
\includegraphics[width=0.6\textwidth]{pandora_plausibility_ratings.jpg}
\end{figure}

@@

As discussed in detail below, the second session showed an adaptation effect \citep{prasad2021rapid}: Reading times decreased overall. The adaptation effect is theoretically interesting per se \citep{fine2013rapid} and we discuss it below; but it was problematic for our design because the average effect would then be confounded by adaptation. For this reason, we ran a second version of the SPR study, hereafter Experiment 1b, in which participants (N=570) were either shown items 1-60 or items 61-120 (but not both). In the Results section, we present both sets of results as well as the pooled data (N=774) from E1a and E1b.
As mentioned above, because the ERP method requires a large number of items, all the ERP data come from participants who completed both sessions.

@@

\begin{figure}[ht]
    \caption{The structure of the present study. In total, 120 items were used for the SPR (red branches) and the ERP experiments (blue branches). The color brightness reflects how many sessions were completed by each participant (lighter color: two sessions, darker color: one session).}
    \label{fig:project_str}
    \centering
    \includegraphics[width=\textwidth]{pandora_project_structure_figure.pdf}
\end{figure}

\section{Data availability statement}
{All materials, data and analysis scripts can be accessed via \\ https://osf.io/4fpru/?view$\_$only=2d2a004d3e4147bc975b5e8c455c27df.}
